---
title: "CA - S7: Non Personalized Recommendation Engine"
author: Josep Curto, IE Business School
abstract: "This document introduces how tobBuilding a non personalized recommendation Engine with R"
keywords: "r, association analysis"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_notebook: 
    fig_caption: yes
    toc: yes
    toc_float: yes
    self_contained: yes
---

# Calculate Non Personalized recommendations with R

## Load packages

```{r packages, warning=FALSE, echo=FALSE, message=FALSE}
# Cleaning the environment
rm(list=ls())

# List of packages for session
.packages <- c("data.table")

# Install CRAN packages (if not already installed)
.inst <- .packages %in% installed.packages()
if(length(.packages[!.inst]) > 0) install.packages(.packages[!.inst])

# Load packages into session 
suppressPackageStartupMessages(invisible(lapply(.packages, library, character.only = TRUE)))
```

## Loading data

First we load the data:

```{r load data}
# Load data
df <- read.csv("data/critics.csv",sep=',', header=T)
```

The summary give is information about the attributes:

```{r summary}
summary(df)
```

Prepare the dataset:

```{r}
rownames(df) <- df$User 
df <- df[,-1]
```

## Top 5 Movies, by rating mean

Calculating mean rating for each movie, ordered with the highest rating listed first, and plot the top 5.

```{r fig.width=7, fig.height=6}
top_rated_movies <- data.table()
for (i in 1:ncol(df)){
  # row binds the result to the existing data table 
  top_rated_movies <- rbind(top_rated_movies, data.table(
                 movie_name = colnames(df)[i],
                 mean_movie = mean(df[,i], na.rm = TRUE)));
}
top_rated_movies[order(-mean_movie)][1:5]
```

## Top 5 Movies, by rating distribution

Calculating the percentage of ratings for each movie that are 4 stars or higher. Order with the highest percentage first, and plot the top 5 movies with more high ratings distribution.

```{r}
top_rated_movies2 <- data.table()
for (i in 1:ncol(df)){
  movie <- df[,i]
  # total number of ratings
  total = sum(movie >= 0, na.rm = T)
  # number of ratings higher or equal 4
  count = sum(movie >= 4, na.rm = T)
  percentage = count / total
  top_rated_movies2 <- rbind(top_rated_movies2, data.table(
                 movie_name = colnames(df)[i],
                 percentage = percentage));
  }
top_rated_movies2[order(-percentage)][1:5]
```

## Top 5 Movies, by quantity of ratings

Counting the number of ratings for each movie, order the one with most number of ratings first, submit the top 5.

```{r}
top_rated_movies3 <- data.table()
for (i in 1:ncol(df)){
  movie <- df[,i]
  total_ratings <- sum(movie >= 0, na.rm = T)
  top_rated_movies3 <- rbind(top_rated_movies3, data.table(
                 movie_name = colnames(df)[i],
                 total_ratings = total_ratings));
  }
top_rated_movies3[order(-total_ratings)][1:5]
```

## People who liked Babe also liked...

Calculate the movies better rated of people who liked a movie. Select the people who liked the movie "Babe" (4 or 5 stars) and provide the top 5 movies they liked most.

```{r}
# number of total ratings of Babe
Babe_count  <-  sum((df[,20] >= 0), na.rm = T) 
# Filter first the rows with a rating for the default movie, filte than the ratings higher equal to four and delete the column of the default movie
df_Babe_movie <- df[!is.na(df$Babe) & (df$Babe>=4) ,][,-20]
top_liked <- data.table()
for (i in 1:ncol(df_Babe_movie)){
  movie <- df_Babe_movie[,i]
  like_mean <-  mean(movie, na.rm = TRUE)
  top_liked <- rbind(top_liked, data.table(
                               movie_name = colnames(df_Babe_movie)[i],
                               like_mean = like_mean));
}
top_liked[order(-like_mean)][1:5]
```
